---
title: "Multivariate Data Analysis: IRIS Dataset"
subtitle: "Principal Component Analysis with R"
author: "Rico Derks, Aswin Verhoeven"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(dplyr)
library(plotly)
library(corrplot)
library(factoextra)
library(FactoMineR)
library(knitr)
library(DT)
library(tidyr)

# Set up tutorial options
tutorial_options(exercise.timelimit = 60)

# Load and prepare the iris data
data(iris)

# Add ID column to match original exercise
iris_data <- iris |>
  mutate(ID = row_number()) |>
  select(ID, Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)

# For exercises without setosa.
# droplevels() removes the now unused "setosa" level, otherwise it keeps
# showing up as an empty entry in the plot legends.
iris_no_setosa <- iris_data |>
  filter(Species != "setosa") |>
  droplevels()
```

## Background

In statistics there are a number of classical data sets which are used to test different methods and algorithms. IRIS is one of these historical data sets and has its origins in botany.

### Objective

In this tutorial you will:

- Create overview models of a full dataset using Principal Component Analysis (PCA)
- Learn to interpret loading plots in terms of variables that contribute to patterns in score plots
- Understand the relationship between scores and loadings in multivariate analysis

### The Data

The dataset contains petal (NL: bloemblad) and sepal (NL: kelkblad) lengths and widths of 50 specimens each of *Iris setosa*, *Iris versicolor*, and *Iris virginica*. The great statistician Fisher introduced this dataset as early as 1936. It is commonly known as "The Fisher Iris Data".

The data is already present in R, so we don't need to specifically load the data. Let's start by examining the data by showing the first 10 rows and calculating some basic statistics:

```{r explore-data1, echo=TRUE}
# Display the iris dataset
datatable(data = iris_data,
          options = list(
            dom = "tip"
          ))

# Get basic statistics
summary(iris_data[, 3:6])  # Only numeric columns
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction technique that finds the directions (principal components) that explain the maximum variance in the data.

### Performing PCA

```{r pca-basic, exercise=TRUE}
# Perform PCA on the numeric variables only
# We'll use the scaled data (equivalent to UV scaling in SIMCA)
pca_result <- prcomp(x = iris_data[, 3:6],
                     scale. = TRUE, 
                     center = TRUE)

# Display the summary
summary(pca_result)
```

```{r pca-basic-solution}
pca_result <- prcomp(iris_data[, 3:6], scale. = TRUE, center = TRUE)
summary(pca_result)
```

### Understanding PCA Results

Let's examine what this summary tells us:

```{r interpret-pca, echo=FALSE}
pca_result <- prcomp(iris_data[,3:6], scale. = TRUE, center = TRUE)
summary_pca <- summary(pca_result)

# Calculate cumulative R2 equivalent
variance_explained <- pca_result$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("Variance Explained by Each Component:\n")
cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("\nCumulative Variance Explained:\n")
cat("PC1:", round(r2_cumulative[1]*100, 1), "%\n")
cat("PC1+PC2:", round(r2_cumulative[2]*100, 1), "%\n")
```

### Quiz: Understanding PCA Quality

```{r quiz-pca-quality, echo=FALSE}
quiz(
  question("Based on the PCA summary, how much variance is explained by the first two principal components combined?",
    answer("About 73%"),
    answer("About 96%", correct = TRUE),
    answer("About 52%"),
    answer("About 23%"),
    message = "The first two components explain approximately 96% of the total variance, indicating excellent model quality."
  ),
  question("If you forced the analysis to calculate more principal components, what's the maximum number it could calculate and why?",
    answer("3 components, because there are 3 species"),
    answer("4 components, because there are 4 variables", correct = TRUE),
    answer("150 components, because there are 150 observations"),
    answer("Unlimited components"),
    message = "PCA can calculate at most min(n-1, p) components, where n is observations and p is variables. With 4 variables, we can have maximum 4 components."
  )
)
```

## Visualizing PCA Results

### Scores Plot

The scores plot shows how the observations (individual iris specimens) are positioned in the new principal component space:

```{r scores-plot, exercise=TRUE, exercise.setup='pca-basic'}
# Create a scores plot
scores_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Species = iris_data$Species,
  ID = iris_data$ID
)

# Calculate cumulative R2 equivalent
variance_explained <- pca_result$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

# Basic ggplot scores plot
ggplot(scores_df, aes(x = PC1,
                      y = PC2, 
                      color = Species)) +
  geom_point(size = 3, 
             alpha = 0.7) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r scores-plot-solution}
scores_df <- data.frame(
  PC1 = pca_result$x[,1],
  PC2 = pca_result$x[,2],
  Species = iris_data$Species,
  ID = iris_data$ID
)

ggplot(scores_df, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Interactive Scores Plot

```{r interactive-scores, exercise=TRUE, exercise.setup="scores-plot"}
# Create an interactive plot with plotly
p <- ggplot(scores_df, aes(x = PC1, 
                           y = PC2, 
                           color = Species,
                           text = paste("ID:", ID))) +
  geom_point(size = 3, 
             alpha = 0.7) +
  labs(title = "Interactive PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)")) +
  theme_minimal()

ggplotly(p, tooltip = c("text", "colour"))
```

### Quiz: Interpreting Scores Plot

```{r quiz-scores, echo=FALSE}
quiz(
  question("What does the scores plot show?",
    answer("The correlation between variables"),
    answer("The position of each observation in the PC space", correct = TRUE),
    answer("The importance of each variable"),
    answer("The raw data values"),
    message = "The scores plot shows how each individual observation (iris specimen) is positioned in the new principal component coordinate system."
  ),
  question("How many distinct clusters can you identify in the scores plot?",
    answer("2 clusters"),
    answer("3 clusters", correct = TRUE),
    answer("1 cluster"),
    answer("No clear clusters"),
    message = "There are 3 distinct clusters corresponding to the 3 iris species, with setosa being most clearly separated."
  )
)
```

## Loadings Plot

The loadings plot shows how the original variables contribute to each principal component:

```{r loadings-plot, exercise=TRUE, exercise.setup='pca-basic'}
# Create loadings plot
loadings_df <- data.frame(
  Variable = rownames(pca_result$rotation),
  PC1 = pca_result$rotation[, 1],
  PC2 = pca_result$rotation[, 2]
)

# Loadings plot
ggplot(data = loadings_df, 
       aes(x = PC1,
           y = PC2)) +
  geom_point(size = 4, 
             color = "red", 
             alpha = 0.7) +
  geom_text(aes(label = Variable), 
            vjust = -0.5,
            hjust = 0.5, 
            size = 3) +
  geom_hline(yintercept = 0, 
             linetype = "dashed",
             alpha = 0.5) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             alpha = 0.5) +
  labs(title = "PCA Loadings Plot",
       x = "PC1 Loadings",
       y = "PC2 Loadings") +
  theme_minimal() +
  xlim(-1, 1) +
  ylim(-1, 1)
```

### Enhanced Loadings Visualization

```{r loadings-arrows, exercise=TRUE, exercise.setup='pca-basic'}
# Create a biplot-style visualization
# Use fviz_pca_biplot for a comprehensive view
fviz_pca_biplot(pca_result, 
                col.ind = iris_data$Species,
                palette = c("#00AFBB", "#E7B800", "#FC4E07"),
                addEllipses = TRUE,
                label = "var",
                col.var = "black",
                repel = TRUE,
                legend.title = "Species") +
  theme_minimal()
```

### Quiz: Interpreting Loadings

```{r quiz-loadings, echo=FALSE}
quiz(
  question("What does the loadings plot show?",
    answer("The position of observations in PC space"),
    answer("How original variables contribute to each PC", correct = TRUE),
    answer("The correlation between species"),
    answer("The raw measurement values"),
    message = "The loadings plot shows how each original variable (Sepal.Length, Sepal.Width, etc.) contributes to the principal components."
  ),
  question("Based on the loadings plot, which variables are most important for separating the clusters in the scores plot?",
    answer("Sepal measurements only"),
    answer("Petal measurements only"),
    answer("Petal.Length and Petal.Width", correct = TRUE),
    answer("All variables equally"),
    message = "Petal.Length and Petal.Width have the strongest loadings on PC1, which explains the most variance and separates the species most clearly."
  )
)
```

## Analysis Without Setosa

Let's examine what happens when we remove the clearly separated setosa species:

```{r pca-no-setosa, exercise=TRUE, exercise.setup='pca-basic'}
# PCA without setosa
pca_no_setosa <- prcomp(iris_no_setosa[, 3:6],
                        scale. = TRUE,
                        center = TRUE)

# Summary
summary(pca_no_setosa)

# Calculate R2 values
variance_no_setosa <- pca_no_setosa$sdev^2
total_variance_no_setosa <- sum(variance_no_setosa)
r2_values_no_setosa <- variance_no_setosa / total_variance_no_setosa
```

```{r plot-no-setosa, exercise=TRUE, exercise.setup="pca-no-setosa"}
# Scores plot without setosa
scores_no_setosa <- data.frame(
  PC1 = pca_no_setosa$x[,1],
  PC2 = pca_no_setosa$x[,2],
  Species = iris_no_setosa$Species
)

ggplot(scores_no_setosa, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA Scores Plot (Without Setosa)",
       x = paste0("PC1 (", round(r2_values_no_setosa[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values_no_setosa[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Comparing Models

```{r compare-models, exercise=TRUE, exercise.setup='pca-no-setosa'}
# Compare the loadings between full data and without setosa
loadings_no_setosa <- data.frame(
  Variable = rownames(pca_no_setosa$rotation),
  PC1 = pca_no_setosa$rotation[,1],
  PC2 = pca_no_setosa$rotation[,2]
)

print("Loadings with all species:")
print(round(pca_result$rotation[,1:2], 3))

print("\nLoadings without setosa:")
print(round(pca_no_setosa$rotation[,1:2], 3))
```

### Quiz: Model Comparison

```{r quiz-comparison, echo=FALSE}
quiz(
  question("When setosa is removed, what happens to the separation between versicolor and virginica?",
    answer("It becomes much clearer"),
    answer("It becomes somewhat clearer", correct = TRUE),
    answer("No change"),
    answer("It becomes less clear"),
    message = "Removing setosa does help reveal the more subtle differences between versicolor and virginica, though they remain more similar to each other than to setosa."
  ),
  question("What is the main variable distinguishing versicolor from virginica?",
    answer("Sepal.Width"),
    answer("Sepal.Length"),
    answer("Petal.Length", correct = TRUE),
    answer("They cannot be distinguished"),
    message = "Looking at the loadings and scores plots, Petal.Length appears to be the primary variable separating these two species along PC1."
  )
)
```

## Alternative Approaches

If multivariate methods don't provide clear separation, what univariate approaches could you use?

```{r univariate-analysis, exercise=TRUE}
# Boxplots for each variable
iris_long <- iris_data |>
  pivot_longer(cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
               names_to = "Variable", values_to = "Value")

ggplot(iris_long, aes(x = Species, y = Value, fill = Species)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Variable, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Each Variable by Species")
```

```{r correlation-analysis, exercise=TRUE}
# Correlation analysis
# Calculate correlation matrix
cor_matrix <- cor(iris_data[,3:6])

# Visualize correlations
corrplot(cor_matrix, method = "circle", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black")
```

## Summary and Key Takeaways

### What We've Learned

1. **PCA Fundamentals**: PCA finds the directions of maximum variance in multivariate data
2. **Scores vs. Loadings**: 
   - Scores show where observations lie in PC space
   - Loadings show how original variables contribute to PCs
3. **Model Quality**: High R² values (>90%) indicate good model fit
4. **Biological Interpretation**: Petal measurements are key distinguishing features of iris species

### Practice Exercise

```{r final-exercise, exercise=TRUE}
# Your turn: Create a comprehensive analysis
# 1. Perform PCA on the full dataset
# 2. Create both scores and loadings plots
# 3. Identify which variables are most important for species separation

# Start your analysis here:
pca_final <- prcomp(iris_data[,3:6], scale. = TRUE)

# Add your plotting code below:

```

```{r final-exercise-solution}
# Complete analysis
pca_final <- prcomp(iris_data[,3:6], scale. = TRUE)

# Scores plot
scores_final <- data.frame(
  PC1 = pca_final$x[, 1],
  PC2 = pca_final$x[, 2],
  Species = iris_data$Species
)

p1 <- ggplot(data = scores_final, 
             aes(x = PC1, 
                 y = PC2, 
                 color = Species)) +
  geom_point(size = 3, 
             alpha = 0.7) +
  labs(title = "Final PCA Analysis - Scores Plot") +
  theme_minimal()

# Loadings information
loadings_final <- pca_final$rotation[,1:2]
print("Most important variables for PC1:")
print(sort(abs(loadings_final[,1]), decreasing = TRUE))

p1
```

### Next Steps

In the next tutorials, we'll explore:

- **Food Dataset**: Geographic patterns in European food consumption
- **NMR Metabolomics**: Advanced PCA and PLS-DA techniques
- **Model validation** and **cross-validation** approaches

Great job completing the IRIS dataset tutorial! You now have a solid foundation in PCA and multivariate data interpretation.