---
title: "Multivariate Data Analysis: IRIS Dataset"
subtitle: "Principal Component Analysis with R"
author: "Rico Derks, Aswin Verhoeven"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "fos-2026-iris"
  version: "1.0"
---

```{r setup, include=FALSE}
library(learnr)
# pcaMethods is loaded first on purpose: it masks a few base/stats functions,
# and loading the tidyverse packages afterwards restores the ones we need.
library(pcaMethods)
library(ggplot2)
library(dplyr)
library(plotly)
library(corrplot)
library(factoextra)
library(knitr)
library(DT)
library(tidyr)

# Set up tutorial options
tutorial_options(exercise.timelimit = 60)

# Load and prepare the iris data
data(iris)

# Add ID column to match original exercise
iris_data <- iris |>
  mutate(ID = row_number()) |>
  select(ID, Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)

# For exercises without setosa.
# droplevels() removes the now unused "setosa" level, otherwise it keeps
# showing up as an empty entry in the plot legends.
iris_no_setosa <- iris_data |>
  filter(Species != "setosa") |>
  droplevels()
```

## How to use this tutorial

Everything happens in your browser - you do not need R installed, and you cannot
break anything. Nothing you type here touches your own computer or the data
files.

**The grey boxes are live R code.** Press **Run Code** to execute the code in a
box and see the output appear underneath it. You are encouraged to change the
code and run it again: change a colour, plot PC2 against PC3 instead of PC1
against PC2, look at a different variable. Experimenting is the point.

- **Run Code** - runs whatever is currently in the box
- **Start Over** - puts the original code back if you get stuck
- **Solution** - some exercises have this button, which shows one possible answer

**The questions** are there to make you look at the output you just produced.
Pick an answer and press **Submit Answer**; you will get an explanation either
way, and the explanation is usually the most useful part. Read it even when you
answered correctly.

**Working through the tutorial.** Sections appear one at a time. Press
**Continue** at the bottom of a section to move on. Your progress is remembered,
so you can close the page and come back later. If you want a completely clean
start, use the **Start Over** link at the bottom of the left-hand menu - note
that this clears the whole tutorial, not just one exercise.

One thing to keep in mind throughout: the code boxes run independently of each
other. An object you create in one box is not automatically available in the
next one, so each exercise sets up what it needs.

## Background

In statistics there are a number of classical data sets which are used to test different methods and algorithms. IRIS is one of these historical data sets and has its origins in botany.

### Objective

In this tutorial you will:

- Create overview models of a full dataset using Principal Component Analysis (PCA)
- Learn to interpret loading plots in terms of variables that contribute to patterns in score plots
- Understand the relationship between scores and loadings in multivariate analysis

### The Data

The dataset contains petal (NL: bloemblad) and sepal (NL: kelkblad) lengths and widths of 50 specimens each of *Iris setosa*, *Iris versicolor*, and *Iris virginica*. The great statistician Fisher introduced this dataset as early as 1936. It is commonly known as "The Fisher Iris Data".

The data is already present in R, so we don't need to specifically load the data. Let's start by examining the data by showing the first 10 rows and calculating some basic statistics:

```{r explore-data1, echo=TRUE}
# Display the iris dataset
datatable(data = iris_data,
          options = list(
            dom = "tip"
          ))

# Get basic statistics
summary(iris_data[, 3:6])  # Only numeric columns
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction technique that finds the directions (principal components) that explain the maximum variance in the data.

### Performing PCA

```{r pca-basic, exercise=TRUE}
# Perform PCA on the numeric variables only
# We'll use the scaled data
pca_result <- prcomp(x = iris_data[, 3:6],
                     scale. = TRUE, 
                     center = TRUE)

# Display the summary
summary(pca_result)
```

```{r pca-basic-solution}
pca_result <- prcomp(iris_data[, 3:6], scale. = TRUE, center = TRUE)
summary(pca_result)
```

### Understanding PCA Results

Let's examine what this summary tells us:

```{r interpret-pca, echo=FALSE}
pca_result <- prcomp(iris_data[,3:6], scale. = TRUE, center = TRUE)
summary_pca <- summary(pca_result)

# Calculate cumulative R2 equivalent
variance_explained <- pca_result$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("Variance Explained by Each Component:\n")
cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("\nCumulative Variance Explained:\n")
cat("PC1:", round(r2_cumulative[1]*100, 1), "%\n")
cat("PC1+PC2:", round(r2_cumulative[2]*100, 1), "%\n")
```

### Model quality: R² and Q²

To be able to say something about the quality of the model it is good to also 
calculate the **Q²** per component as well:

- **R²cum** - the cumulative fraction of the data *described* by the model.
  This is simply the cumulative variance explained that we calculated above.
  R² can only go up when you add components, so on its own it says very little.
- **Q²cum** - the cumulative fraction of the data that the model can
  *predict*. Q² is obtained by cross-validation: the observations are split
  into 7 groups, one group is left out, a new model is built on the remaining
  6/7, and the left-out observations are then predicted. This is repeated until
  every observation has been left out once. The closer the predictions are to
  the real values, the higher Q².

The important difference: **R² always increases when you add a component, but
Q² does not**. Once a component starts describing noise instead of structure,
Q² stops rising and begins to fall. 

R does not report Q² by default, but the `pcaMethods` package can calculate it:

```{r r2-q2, exercise=TRUE, exercise.setup='pca-basic'}
# pcaMethods has its own pca() function which can report both R2 and Q2
set.seed(2026)  # cross-validation splits the data randomly

X <- as.matrix(iris_data[, 3:6])

pca_cv <- pca(X,
              method = "svd",
              nPcs = 4,
              scale = "uv",     # unit variance, same as scale. = TRUE
              center = TRUE)

# R2cum: cumulative variance described
r2cum <- R2cum(pca_cv)

# Q2: cumulative predictive ability, 7-fold cross-validation
q2cum <- Q2(pca_cv, X, fold = 7, verbose = FALSE)

data.frame(
  Component = paste0("PC", seq_along(r2cum)),
  R2cum = round(r2cum, 3),
  Q2cum = c(round(q2cum, 3), rep(NA, length(r2cum) - length(q2cum)))
)
```

```{r r2-q2-solution}
set.seed(2026)
X <- as.matrix(iris_data[, 3:6])
pca_cv <- pca(X, method = "svd", nPcs = 4, scale = "uv", center = TRUE)
r2cum <- R2cum(pca_cv)
q2cum <- Q2(pca_cv, X, fold = 7, verbose = FALSE)

data.frame(
  Component = paste0("PC", seq_along(r2cum)),
  R2cum = round(r2cum, 3),
  Q2cum = c(round(q2cum, 3), rep(NA, length(r2cum) - length(q2cum)))
)
```

### Quiz: Understanding PCA Quality

```{r quiz-pca-quality, echo=FALSE}
quiz(
  question("Based on the PCA summary, how much variance is explained by the first two principal components combined?",
    answer("About 73%"),
    answer("About 96%", correct = TRUE),
    answer("About 52%"),
    answer("About 23%"),
    message = "The first two components explain approximately 96% of the total variance, indicating excellent model quality."
  ),
  question("If you forced the analysis to calculate more principal components, what's the maximum number it could calculate and why?",
    answer("3 components, because there are 3 species"),
    answer("4 components, because there are 4 variables", correct = TRUE),
    answer("150 components, because there are 150 observations"),
    answer("Unlimited components"),
    message = "PCA can calculate at most min(n-1, p) components, where n is observations and p is variables. With 4 variables, we can have maximum 4 components. Notice in the table above that R2cum reaches 1.000 at the fourth component: the model then describes the data completely and there is nothing left to model."
  ),
  question("The model has R2cum = 0.96 and Q2cum = 0.71 after two components. What does this tell you?",
    answer("Two components describe 96% of the data and also predict it well: a good model", correct = TRUE),
    answer("The model is overfitted, because R2 is much higher than Q2"),
    answer("The model is useless, because Q2 is below 1"),
    answer("Nothing, R2 and Q2 cannot be compared"),
    message = "R2 tells you how much is described, Q2 how much can be predicted. A high R2 with a high Q2 means the model captures real structure rather than noise. In metabolomics a Q2 of around 0.5 is already considered good, so 0.71 is high. R2 is always larger than Q2; only a LARGE gap between them is a warning sign.",
    correct = "Correct. Note that R2 and Q2 together still say nothing about outliers - we look at those next."
  )
)
```

## Visualizing PCA Results

### Scores Plot

The scores plot shows how the observations (individual iris specimens) are positioned in the new principal component space:

```{r scores-plot, exercise=TRUE, exercise.setup='pca-basic', fig.width=7, fig.height=6}
# Create a scores plot
scores_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Species = iris_data$Species,
  ID = iris_data$ID
)

# Calculate cumulative R2 equivalent
variance_explained <- pca_result$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

# Basic ggplot scores plot
p <- ggplot(scores_df, aes(x = PC1,
                           y = PC2, 
                           color = Species)) +
  geom_vline(xintercept = 0,
             color = "grey") +
  geom_hline(yintercept = 0,
             color = "grey") +
  geom_point(size = 3, 
             alpha = 0.7) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = c("text", "colour", "x", "y"))
```

```{r scores-plot-solution}
scores_df <- data.frame(
  PC1 = pca_result$x[,1],
  PC2 = pca_result$x[,2],
  Species = iris_data$Species,
  ID = iris_data$ID
)

ggplot(scores_df, aes(x = PC1, y = PC2, color = Species)) +
  geom_vline(xintercept = 0,
             color = "grey") +
  geom_hline(yintercept = 0,
             color = "grey") +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = c("text", "colour", "x", "y"))
```

### Quiz: Interpreting Scores Plot

```{r quiz-scores, echo=FALSE}
quiz(
  question("What does the scores plot show?",
    answer("The correlation between variables"),
    answer("The position of each observation in the PC space", correct = TRUE),
    answer("The importance of each variable"),
    answer("The raw data values"),
    message = "The scores plot shows how each individual observation (iris specimen) is positioned in the new principal component coordinate system."
  ),
  question("How many distinct clusters can you identify in the scores plot?",
    answer("2 clusters"),
    answer("3 clusters", correct = TRUE),
    answer("1 cluster"),
    answer("No clear clusters"),
    message = "There are 3 distinct clusters corresponding to the 3 iris species, with setosa being most clearly separated."
  )
)
```

## Outliers: Hotelling's T² and DModX

R² and Q² tell you how much of the data the model describes and predicts, but
they say nothing about *individual observations*. A single strange sample can
distort a model without making R² or Q² look bad. Therefore, it is important to 
look at two diagnostic plots.

**Hotelling's T²** finds *strong* outliers: observations that lie far from the
centre **within** the model plane. In a scores plot this is drawn as a
confidence ellipse - anything outside it is a candidate strong outlier.

```{r hotelling, exercise=TRUE, exercise.setup='pca-basic', fig.width=7, fig.height=6}
scores_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Species = iris_data$Species,
  ID = iris_data$ID
)

# Hotelling's T2 for the first two components
n <- nrow(scores_df)
A <- 2 # number of components
sd_pc <- apply(pca_result$x[, 1:A], 2, sd)

scores_df$T2 <- rowSums((pca_result$x[, 1:A] / rep(sd_pc, each = n))^2)

# 95% critical value
t2_crit <- (A * (n - 1) * (n + 1)) / (n * (n - A)) * qf(0.95, A, n - A)

# The ellipse itself: semi-axes are sd * sqrt(critical value)
theta <- seq(0, 2 * pi, length.out = 200)
ellipse <- data.frame(
  PC1 = sd_pc[1] * sqrt(t2_crit) * cos(theta),
  PC2 = sd_pc[2] * sqrt(t2_crit) * sin(theta)
)

p <- ggplot(scores_df, aes(x = PC1, y = PC2)) +
  geom_vline(xintercept = 0,
             color = "grey") +
  geom_hline(yintercept = 0,
             color = "grey") +
  geom_path(data = ellipse, colour = "grey40", linetype = "dashed") +
  geom_point(aes(colour = Species), size = 3, alpha = 0.7) +
  geom_text(data = subset(scores_df, T2 > t2_crit),
            aes(label = ID), vjust = -0.9, size = 3) +
  labs(title = "Scores plot with 95% Hotelling's T² ellipse") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = c("text", "colour", "x", "y"))

cat("Observations outside the ellipse:", sum(scores_df$T2 > t2_crit), "of", n, "\n")
```

**DModX** (Distance to the Model in X) finds *moderate* outliers: observations
that are not extreme within the plane, but lie far *away* from it (rule-of-thumb 
2 - 3x), meaning the model describes them poorly.

```{r dmodx, exercise=TRUE, exercise.setup='pca-basic', fig.width=7, fig.height=6}
X <- as.matrix(iris_data[, 3:6])

pca_cv <- pca(X, method = "svd", nPcs = 2, scale = "uv", center = TRUE)

dmodx <- DModX(pca_cv, X)

dmodx_df <- data.frame(
  ID = iris_data$ID,
  Species = iris_data$Species,
  DModX = as.numeric(dmodx)
)

# The normalised DModX is scaled so that 1 is the average distance to the
# model. A common rule of thumb is that an observation is only worth calling
# an outlier once it is about twice that.
p <- ggplot(dmodx_df, aes(x = ID, y = DModX, fill = Species)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_hline(yintercept = 2, linetype = "dotted", colour = "red") +
  labs(y = "DModX (normalised)") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = c("text", "colour", "x", "y")) |> 
  layout(
    title = list(
      text = paste0(
        "DModX per observation",
        "<br>",
        "<sup>Dashed = model average (1), red dotted = rule-of-thumb limit (2)</sup>"
      ),
      x = 0,
      y = 1,
      xanchor = "left",
      yanchor = "top",
      xref = "paper",
      yref = "paper"
    )
  )

# Which observations are furthest from the model?
head(dmodx_df[order(-dmodx_df$DModX), ], 5)
```

### Quiz: Outlier diagnostics

```{r quiz-outliers, echo=FALSE}
quiz(
  question("What is the difference between Hotelling's T² and DModX?",
    answer("T² finds observations far from the centre within the model plane; DModX finds observations far away from the plane", correct = TRUE),
    answer("They measure exactly the same thing"),
    answer("T² is for variables, DModX is for observations"),
    answer("T² is calculated on the loadings, DModX on the scores"),
    message = "T² measures distance from the origin WITHIN the model (strong outliers, visible in the scores plot). An observation far away from the center can have a high leverage. DModX measures the residual, i.e. the distance from the model plane (moderate outliers, invisible in a scores plot) High DModX for an observation means that the model describes this observation poorly. You need both!"
  ),
  question("A few observations lie just outside the 95% T² ellipse. Should you remove them?",
    answer("No - investigate them first; being outside a 95% ellipse is expected for about 5% of observations", correct = TRUE),
    answer("Yes, always remove anything outside the ellipse"),
    answer("Yes, but only if they belong to setosa"),
    answer("It makes no difference either way"),
    message = "By definition roughly 5% of the observations fall outside a 95% ellipse even when nothing is wrong - with 150 flowers you expect about 7, and here there are 6. A common rule of thumb for DModX is that a value needs to be about twice the model average before you call it an outlier, and even then you should check the raw data first. Removing observations just because they are extreme is a good way to fool yourself."
  )
)
```

## Loadings Plot

The loadings plot shows how the original variables contribute to each principal component:

<div class="alert alert-info">
**A word of warning about signs.** The direction of a principal component is
arbitrary. A component and its mirror image describe exactly the same model, so
different software - or even the same software on a different computer - may
flip PC1 or PC2 upside down. 

What is *not* arbitrary is the **relative** position of things: which variables
sit together, which sit on opposite sides of the origin, and how far they are
from the centre. Always interpret loadings relative to each other, never as
"positive is good, negative is bad".
</div>

```{r loadings-plot, exercise=TRUE, exercise.setup='pca-basic', fig.width=7, fig.height=6}
# Create loadings plot
loadings_df <- data.frame(
  Variable = rownames(pca_result$rotation),
  PC1 = pca_result$rotation[, 1],
  PC2 = pca_result$rotation[, 2]
)

# Loadings plot
p <- ggplot(data = loadings_df, 
            aes(x = PC1,
                y = PC2)) +
  geom_point(size = 4, 
             color = "red", 
             alpha = 0.7) +
  geom_text(aes(label = Variable), 
            vjust = -0.5,
            hjust = 0.5, 
            size = 3) +
  geom_hline(yintercept = 0, 
             linetype = "dashed",
             alpha = 0.5) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             alpha = 0.5) +
  labs(title = "PCA Loadings Plot",
       x = "PC1 Loadings",
       y = "PC2 Loadings") +
  theme_minimal() +
  xlim(-1, 1) +
  ylim(-1, 1)

ggplotly(p, tooltip = c("text", "colour", "x", "y"))
```

### Enhanced Loadings Visualization

```{r loadings-arrows, exercise=TRUE, exercise.setup='pca-basic', fig.width=7, fig.height=6}
# Create a biplot-style visualization
# Use fviz_pca_biplot for a comprehensive view
fviz_pca_biplot(pca_result, 
                col.ind = iris_data$Species,
                palette = c("#00AFBB", "#E7B800", "#FC4E07"),
                addEllipses = TRUE,
                label = "var",
                col.var = "black",
                repel = TRUE,
                legend.title = "Species") +
  labs(x = "PC1 Loadings",
       y = "PC2 Loadings") +
  theme_minimal()
```

### Quiz: Interpreting Loadings

```{r quiz-loadings, echo=FALSE}
quiz(
  question("What does the loadings plot show?",
    answer("The position of observations in PC space"),
    answer("How original variables contribute to each PC", correct = TRUE),
    answer("The correlation between species"),
    answer("The raw measurement values"),
    message = "The loadings plot shows how each original variable (Sepal.Length, Sepal.Width, etc.) contributes to the principal components."
  ),
  question("Based on the loadings plot, which variables are most important for separating the clusters in the scores plot?",
    answer("Petal.Length and Petal.Width", correct = TRUE),
    answer("Sepal.Length and Sepal.Width"),
    answer("Sepal.Width on its own"),
    answer("All four variables equally"),
    message = "PC1 carries 73% of the variance and does most of the separating. Its loadings are Petal.Length 0.58, Petal.Width 0.57, Sepal.Length 0.52 and Sepal.Width -0.27. The two petal measurements are the largest, and note that Sepal.Width has the opposite sign: setosa has the smallest petals but the widest sepals of the three species."
  )
)
```

## Analysis Without Setosa

Let's examine what happens when we remove the clearly separated setosa species:

```{r pca-no-setosa, exercise=TRUE, exercise.setup='pca-basic'}
# PCA without setosa
pca_no_setosa <- prcomp(iris_no_setosa[, 3:6],
                        scale. = TRUE,
                        center = TRUE)

# Summary
summary(pca_no_setosa)

# Calculate R2 values
variance_no_setosa <- pca_no_setosa$sdev^2
total_variance_no_setosa <- sum(variance_no_setosa)
r2_values_no_setosa <- variance_no_setosa / total_variance_no_setosa
```

```{r plot-no-setosa, exercise=TRUE, exercise.setup="pca-no-setosa"}
# Scores plot without setosa
scores_no_setosa <- data.frame(
  PC1 = pca_no_setosa$x[,1],
  PC2 = pca_no_setosa$x[,2],
  Species = iris_no_setosa$Species
)

ggplot(scores_no_setosa, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA Scores Plot (Without Setosa)",
       x = paste0("PC1 (", round(r2_values_no_setosa[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values_no_setosa[2]*100, 1), "% variance)")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Comparing Models

```{r compare-models, exercise=TRUE, exercise.setup='pca-no-setosa'}
# Compare the loadings between full data and without setosa
loadings_no_setosa <- data.frame(
  Variable = rownames(pca_no_setosa$rotation),
  PC1 = pca_no_setosa$rotation[,1],
  PC2 = pca_no_setosa$rotation[,2]
)

print("Loadings with all species:")
print(round(pca_result$rotation[,1:2], 3))

print("\nLoadings without setosa:")
print(round(pca_no_setosa$rotation[,1:2], 3))
```

### Quiz: Model Comparison

```{r quiz-comparison, echo=FALSE}
quiz(
  question("When setosa is removed, what happens to the separation between versicolor and virginica?",
    answer("It becomes somewhat clearer, but the two species still overlap", correct = TRUE),
    answer("They separate completely"),
    answer("No change at all"),
    answer("It becomes less clear"),
    message = "Setosa no longer dominates the model, so the axes now describe the remaining two species. But look carefully at the scores plot: there is still substantial overlap between versicolor and virginica. PCA is unsupervised - it looks for the largest variance, not for the difference between your groups, and there is no guarantee those are the same thing."
  ),
  question("Look at the loadings of the model without setosa. How would you best describe PC1?",
    answer("A general 'size' axis: petal length, petal width and sepal length all load strongly in the same direction", correct = TRUE),
    answer("An axis driven by Sepal.Width alone"),
    answer("An axis contrasting petal measurements against sepal measurements"),
    answer("An axis with no sensible interpretation"),
    message = "Without setosa, PC1 covers 73.9% of the variance and all four variables load in the SAME direction, with magnitudes Petal.Length 0.54, Petal.Width 0.51, Sepal.Length 0.51 and Sepal.Width 0.44. (Whether R prints these as positive or negative is arbitrary - only the relative signs matter.) So PC1 is essentially overall flower size, with the three largest contributions coming from petal length, petal width and sepal length. Virginica flowers are simply bigger than versicolor ones, which is why no single variable separates them cleanly."
  )
)
```

## Alternative Approaches

If multivariate methods don't provide clear separation, what univariate approaches could you use?

```{r quiz-univariate, echo=FALSE}
question("Suppose PCA does not separate versicolor from virginica well enough. Which univariate approaches could you use instead? (select all that apply)",
  answer("A t-test per variable, with a correction for multiple testing", correct = TRUE),
  answer("A one-way ANOVA per variable when you have more than two groups", correct = TRUE),
  answer("Simply pick the variable with the largest PC1 loading and report it without further testing"),
  answer("Nothing - if PCA fails there is no alternative"),
  type = "multiple",
  message = "The standard fallback is to go variable by variable: a t-test for two groups, or a one-way ANOVA when there are three or more. Because you then perform one test per variable, you must correct for multiple testing (for example Bonferroni or Benjamini-Hochberg), otherwise you will find 'significant' differences by chance alone. With only 4 variables that is a minor issue, but in the NMR tutorial you will have 655 of them, and there it matters a great deal."
)
```

The boxplots below are the visual version of exactly that univariate comparison:

```{r univariate-analysis, exercise=TRUE}
# Boxplots for each variable
iris_long <- iris_data |>
  pivot_longer(cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width),
               names_to = "Variable", values_to = "Value")

ggplot(iris_long, aes(x = Species, y = Value, fill = Species)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Variable, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Each Variable by Species")
```

```{r correlation-analysis, exercise=TRUE}
# Correlation analysis
# Calculate correlation matrix
cor_matrix <- cor(iris_data[,3:6])

# Visualize correlations
corrplot(cor_matrix, method = "circle", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black")
```

## Summary and Key Takeaways

### What We've Learned

1. **PCA Fundamentals**: PCA finds the directions of maximum variance in multivariate data
2. **Scores vs. Loadings**: 
   - Scores show where observations lie in PC space
   - Loadings show how original variables contribute to PCs
3. **Model Quality**: R² tells you how much is *described*, Q² (from cross-validation) how much can be *predicted*. R² always rises when you add components; Q² does not, and that is what tells you when to stop. Here R²cum = 0.96 and Q²cum = 0.71 after two components.
4. **Diagnostics**: R² and Q² say nothing about individual observations. Hotelling's T² finds strong outliers within the model plane, DModX finds observations lying far from it. Always look at both before trusting a model.
5. **Sign conventions**: the direction of a component is arbitrary; interpret loadings relative to each other, not by their absolute sign.
6. **Biological Interpretation**: Petal measurements are key distinguishing features of iris species

### Practice Exercise

```{r final-exercise, exercise=TRUE}
# Your turn: Create a comprehensive analysis
# 1. Perform PCA on the full dataset
# 2. Create both scores and loadings plots
# 3. Identify which variables are most important for species separation

# Start your analysis here:
pca_final <- prcomp(iris_data[,3:6], scale. = TRUE)

# Add your plotting code below:

```

```{r final-exercise-solution}
# Complete analysis
pca_final <- prcomp(iris_data[,3:6], scale. = TRUE)

# Scores plot
scores_final <- data.frame(
  PC1 = pca_final$x[, 1],
  PC2 = pca_final$x[, 2],
  Species = iris_data$Species
)

p1 <- ggplot(data = scores_final, 
             aes(x = PC1, 
                 y = PC2, 
                 color = Species)) +
  geom_point(size = 3, 
             alpha = 0.7) +
  labs(title = "Final PCA Analysis - Scores Plot") +
  theme_minimal()

# Loadings information
loadings_final <- pca_final$rotation[,1:2]
print("Most important variables for PC1:")
print(sort(abs(loadings_final[,1]), decreasing = TRUE))

p1
```

### Next Steps

In the next tutorials, we'll explore:

- **Food Dataset**: Geographic patterns in European food consumption
- **NMR Metabolomics**: Advanced PCA and PLS-DA techniques
- **Model validation** and **cross-validation** approaches

Great job completing the IRIS dataset tutorial! You now have a solid foundation in PCA and multivariate data interpretation.