---
title: "Multivariate Data Analysis: Food data set"
subtitle: "Principal Component Analysis with R"
author: "Rico Derks, Aswin Verhoeven"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(dplyr)
library(plotly)
library(factoextra)
library(FactoMineR)
library(knitr)
library(DT)
library(tidyr)

# Set up tutorial options
tutorial_options(exercise.timelimit = 60)

# load the data
food_data <- read.csv(
  file = "./foods.csv",
  header = TRUE
)

# Store variable names for easy reference
food_vars <- names(food_data)[5:ncol(food_data)]
n_vars <- length(food_vars)

# Imputed version, used by all exercises from the PCA section onwards.
# (See the "Handling Missing Values" section for how this is done.)
food_imputed <- food_data
for (col in food_vars) {
  food_imputed[[col]][is.na(food_imputed[[col]])] <- mean(food_imputed[[col]],
                                                          na.rm = TRUE)
}

# Colours and labels for the geographic regions, used in several plots
region_colors <- c("C" = "#E69F00", "S" = "#D55E00", "N" = "#0072B2",
                   "U" = "#009E73", "X" = "#CC79A7")
region_labels <- c("C" = "Central", "S" = "Southern", "N" = "Northern",
                   "U" = "UK/Ireland", "X" = "Benelux")
```

```{r prep-pca}
# Setup chunk: makes the PCA model available to the exercises below.
# learnr runs each exercise in its own environment, so anything an exercise
# needs has to come from the global setup chunk or from a setup chunk like
# this one, referenced with the exercise.setup chunk option.
# (Note: a setup chunk must NOT be named "<something>-setup" unless an
# exercise with that exact name exists, so we use the "prep-" prefix.)
pca_food <- prcomp(food_imputed[, food_vars],
                   scale. = TRUE,
                   center = TRUE)

variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)
```

```{r prep-scores, exercise.setup="prep-pca"}
scores_df <- data.frame(
  PC1 = pca_food$x[, 1],
  PC2 = pca_food$x[, 2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location,
  Capitol_Latitude = food_imputed$Capitol_Latitude
)
```

## Background

Data were collected to investigate the consumption pattern of a number of provisions in different European countries. The purpose of the investigation was to examine similarities and differences between the countries and the possible explanations.

### Objective

In this tutorial you will:

- Explain why there are groupings in multivariate plots
- Identify data characteristics that differentiate Portugal and Spain from Sweden and Denmark
- Understand geographic and cultural patterns in food consumption across Europe
- Practice interpreting PCA results with real-world data

### The Data

The dataset consists of **20 variables** (different foods) and **16 observations** (European countries). The values are the percentages of households in each country where a particular product was found.

Let's explore the data structure:

```{r explore-data, exercise=TRUE}
# Display the first few rows
head(food_data, 10)

# Get summary statistics for the food variables
summary(food_data[, food_vars])
```

```{r explore-data-solution}
head(food_data, 10)
summary(food_data[, food_vars])
```

### Data Exploration

```{r data-structure, exercise=TRUE}
# Check the structure
str(food_data)

# Count countries by geographic location
table(food_data$Geographic_Location)

# Geographic locations:
# C = Central, S = Southern, N = Northern, U = UK/Ireland, X = Benelux
```

<div id="data-structure-hint">
**Hint:** The `Geographic_Location` variable groups the countries:
C = Central Europe, S = Southern Europe, N = Northern Europe,
U = UK/Ireland, X = Benelux.
</div>

### Interactive Data Table

```{r data-table, exercise=TRUE}
# Create an interactive table to explore the data
datatable(food_data,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "European Food Consumption Data")
```

## Principal Component Analysis

### Handling Missing Values

Notice that some values are missing (NA). We need to address this before PCA:

```{r check-missing, exercise=TRUE}
# Check for missing values
colSums(is.na(food_data))

# Visualize missing data pattern
missing_pattern <- sapply(food_data[, food_vars], function(x) sum(is.na(x)))
barplot(missing_pattern, las = 2, main = "Missing Values per Variable",
        cex.names = 0.7, col = "steelblue")
```

```{r handle-missing, exercise=TRUE}
# Option 1: Remove rows with missing values (simple approach)
food_complete <- na.omit(food_data)
cat("Countries remaining:", nrow(food_complete), "\n")
cat("Countries removed:", food_data$Country[!complete.cases(food_data)])

# Option 2: Impute missing values with column means
food_imputed <- food_data
for(col in food_vars) {
  food_imputed[[col]][is.na(food_imputed[[col]])] <- mean(food_imputed[[col]], na.rm = TRUE)
}

# We'll use the imputed version for our analysis
cat("\nUsing imputed dataset with all", nrow(food_imputed), "countries")
```

### Performing PCA

```{r pca-basic, exercise=TRUE}
# Perform PCA on the food variables (columns 5 onwards)
# Using scale = TRUE for standardization (UV scaling)
pca_food <- prcomp(food_imputed[, food_vars], 
                   scale. = TRUE, 
                   center = TRUE)

# Display summary
summary(pca_food)

# Calculate variance explained
variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("\nVariance explained by first 3 PCs:\n")
cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("PC3:", round(r2_values[3]*100, 1), "%\n")
cat("\nCumulative variance (PC1+PC2):", round(r2_cumulative[2]*100, 1), "%\n")
cat("Cumulative variance (PC1+PC2+PC3):", round(r2_cumulative[3]*100, 1), "%\n")
```

```{r pca-basic-solution}
pca_food <- prcomp(food_imputed[, food_vars], scale. = TRUE, center = TRUE)
summary(pca_food)

variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("Cumulative (PC1+PC2):", round(r2_cumulative[2]*100, 1), "%\n")
```

### Quiz: Model Quality

```{r quiz-pca-quality, echo=FALSE}
quiz(
  question("Based on the PCA summary, how many principal components would you consider sufficient to describe the data?",
    answer("1 component"),
    answer("2 components", correct = TRUE),
    answer("3 components"),
    answer("All 20 components"),
    message = "The first two components typically explain around 50-60% of the variance, which is reasonable for this type of data. Adding a third component helps capture more nuance."
  ),
  question("What can you say about the quality of this PCA model compared to the IRIS dataset?",
    answer("Much better - higher R²"),
    answer("About the same"),
    answer("Not as good - lower R²", correct = TRUE),
    answer("Cannot compare different datasets"),
    message = "Food consumption data is more complex and variable than flower measurements, so we expect lower R² values. This is normal for behavioral/social data."
  )
)
```

## Scores Plot (PC1 vs PC2)

The scores plot shows how countries are positioned based on their food consumption patterns:

```{r scores-plot, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare scores data
scores_df <- data.frame(
  PC1 = pca_food$x[,1],
  PC2 = pca_food$x[,2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location,
  Capitol_Latitude = food_imputed$Capitol_Latitude
)

# Create scores plot colored by geographic location
ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, hjust = 0.5, size = 3) +
  labs(title = "PCA Scores Plot - European Food Consumption",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r scores-plot-solution}
scores_df <- data.frame(
  PC1 = pca_food$x[,1],
  PC2 = pca_food$x[,2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location
)

ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, size = 3) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "%)")) +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal()
```

### Interactive Scores Plot

```{r interactive-scores, exercise=TRUE, exercise.setup="prep-scores"}
# Create interactive plot with plotly
p <- ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location,
                           text = paste("Country:", Country))) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(aes(label = Country), vjust = -0.7, size = 3, show.legend = FALSE) +
  labs(title = "Interactive PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "%)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal()

ggplotly(p, tooltip = c("text", "colour"))
```

### Quiz: Identifying Clusters

```{r quiz-clusters, echo=FALSE}
quiz(
  question("Are there clear clusters present in the scores plot?",
    answer("No clusters"),
    answer("Yes, some regional patterns", correct = TRUE),
    answer("Perfect separation by region"),
    answer("Only 2 distinct clusters"),
    message = "You can see some regional groupings, particularly Northern European countries clustering together, and Mediterranean countries showing similarity."
  ),
  question("Which countries appear most different from the others?",
    answer("Germany and France"),
    answer("England and Ireland", correct = TRUE),
    answer("Sweden and Norway"),
    answer("Portugal and Spain"),
    message = "England and Ireland (UK region) show distinctive food consumption patterns, separating them from continental Europe."
  )
)
```

## Loadings Plot

The loadings plot shows which food items contribute most to the patterns we see:

```{r loadings-plot, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare loadings data
loadings_df <- data.frame(
  Variable = rownames(pca_food$rotation),
  PC1 = pca_food$rotation[,1],
  PC2 = pca_food$rotation[,2]
)

# Create loadings plot
ggplot(loadings_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, color = "red", alpha = 0.7) +
  geom_text(aes(label = Variable), vjust = -0.5, hjust = 0.5, size = 2.5, check_overlap = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PCA Loadings Plot",
       x = "PC1 Loadings",
       y = "PC2 Loadings") +
  theme_minimal() +
  coord_fixed()
```

### Identify Top Contributing Variables

```{r top-loadings, exercise=TRUE, exercise.setup="prep-pca"}
# Find variables with strongest loadings on PC1
pc1_loadings <- sort(abs(pca_food$rotation[,1]), decreasing = TRUE)
cat("Top 5 variables contributing to PC1:\n")
print(head(pc1_loadings, 5))

# Find variables with strongest loadings on PC2
pc2_loadings <- sort(abs(pca_food$rotation[,2]), decreasing = TRUE)
cat("\nTop 5 variables contributing to PC2:\n")
print(head(pc2_loadings, 5))

# Show actual direction (positive or negative)
cat("\nPC1 loadings (with direction):\n")
print(sort(pca_food$rotation[,1], decreasing = TRUE)[1:5])
cat("\nPC1 loadings (negative direction):\n")
print(sort(pca_food$rotation[,1], decreasing = FALSE)[1:5])
```

### Quiz: Interpreting Loadings

```{r quiz-loadings, echo=FALSE}
quiz(
  question("What food items are causing the main clustering patterns in the scores plot?",
    answer("Coffee and tea only"),
    answer("Garlic and Olive oil", correct = TRUE),
    answer("Butter and margarine only"),
    answer("All foods equally"),
    message = "Garlic and Olive oil have strong loadings and clearly distinguish Mediterranean countries from Northern European countries."
  )
)
```

## PC1 vs PC3 Analysis

Let's examine PC1 vs PC3 to understand other patterns:

```{r scores-pc1-pc3, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare scores data for PC1 vs PC3
scores_df_pc3 <- data.frame(
  PC1 = pca_food$x[,1],
  PC3 = pca_food$x[,3],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location
)

# Create scores plot
ggplot(scores_df_pc3, aes(x = PC1, y = PC3, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, hjust = 0.5, size = 3) +
  labs(title = "PCA Scores Plot - PC1 vs PC3",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC3 (", round(r2_values[3]*100, 1), "%)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### PC3 Loadings

```{r loadings-pc3, exercise=TRUE, exercise.setup="prep-pca"}
# Examine PC3 loadings
loadings_pc3 <- data.frame(
  Variable = rownames(pca_food$rotation),
  PC1 = pca_food$rotation[,1],
  PC3 = pca_food$rotation[,3]
)

ggplot(loadings_pc3, aes(x = PC1, y = PC3)) +
  geom_point(size = 3, color = "red", alpha = 0.7) +
  geom_text(aes(label = Variable), vjust = -0.5, hjust = 0.5, size = 2.5, check_overlap = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PCA Loadings Plot - PC1 vs PC3",
       x = "PC1 Loadings",
       y = "PC3 Loadings") +
  theme_minimal() +
  coord_fixed()

# Show top contributors to PC3
cat("\nTop contributors to PC3:\n")
print(sort(abs(pca_food$rotation[,3]), decreasing = TRUE)[1:5])
```

### Detailed Country Comparison

```{r country-detail, exercise=TRUE}
# Compare specific countries mentioned in the exercise
countries_of_interest <- c("England", "Ireland", "Portugal", "Spain", 
                           "Sweden", "Denmark")

# Extract their data
subset_data <- food_imputed %>%
  filter(Country %in% countries_of_interest) %>%
  select(Country, Garlic, Olive_Oil, Tea, Ground_Coffee)

print(subset_data)

# Visualize
subset_long <- subset_data %>%
  pivot_longer(cols = c(Garlic, Olive_Oil, Tea, Ground_Coffee),
               names_to = "Food", values_to = "Percentage")

ggplot(subset_long, aes(x = Country, y = Percentage, fill = Food)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Key Food Items - Selected Countries",
       y = "Percentage of Households")
```

### Quiz: PC1 vs PC3 Interpretation

```{r quiz-pc3, echo=FALSE}
quiz(
  question("What separates England and Ireland from the rest of Europe on PC1 vs PC3?",
    answer("Their high tea consumption", correct = TRUE),
    answer("Their high coffee consumption"),
    answer("Their butter usage"),
    answer("Their yoghurt consumption"),
    message = "England and Ireland have distinctively high tea consumption (99%) compared to most continental European countries."
  ),
  question("How does the Garlic and Olive oil intake of England/Ireland compare to the rest of Europe?",
    answer("Much higher than average"),
    answer("About average"),
    answer("Much lower than average", correct = TRUE),
    answer("Only olive oil is lower"),
    message = "England and Ireland have very low garlic (2%) and olive oil (1%) consumption compared to Mediterranean countries like Spain (23% garlic, 49% olive oil)."
  )
)
```

## Enhanced Visualizations

### Biplot - Scores and Loadings Together

```{r biplot, exercise=TRUE, exercise.setup="prep-pca"}
# Create a biplot showing both scores and loadings
fviz_pca_biplot(pca_food,
                label = "var",
                col.ind = food_imputed$Geographic_Location,
                palette = region_colors,
                addEllipses = FALSE,
                col.var = "black",
                repel = TRUE,
                legend.title = "Region",
                title = "PCA Biplot - Food Consumption Patterns") +
  theme_minimal()
```

### Contribution of Variables

```{r variable-contribution, exercise=TRUE, exercise.setup="prep-pca"}
# Visualize variable contributions to PC1 and PC2
# Contributions to PC1
fviz_contrib(pca_food, choice = "var", axes = 1, top = 10) +
  labs(title = "Top 10 Variables Contributing to PC1") +
  theme_minimal()
```

```{r variable-contribution-pc2, exercise=TRUE, exercise.setup="prep-pca"}
# Contributions to PC2
fviz_contrib(pca_food, choice = "var", axes = 2, top = 10) +
  labs(title = "Top 10 Variables Contributing to PC2") +
  theme_minimal()
```

## Alternative Univariate Approaches

If multivariate methods don't provide sufficient clarity, what alternative approaches could you use?

```{r univariate-viz, exercise=TRUE}
# Heatmap of food consumption by country
# Select a subset of interesting foods
foods_subset <- c("Garlic", "Olive_Oil", "Tea", "Ground_Coffee", 
                  "Butter", "Margarine", "Frozen_Vegetables")

heatmap_data <- food_imputed %>%
  select(Country, all_of(foods_subset))

# Convert to long format
heatmap_long <- heatmap_data %>%
  pivot_longer(cols = -Country, names_to = "Food", values_to = "Percentage")

# Create heatmap
ggplot(heatmap_long, aes(x = Food, y = Country, fill = Percentage)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", 
                       midpoint = 50) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Food Consumption by Country",
       fill = "% Households")
```

### Clustering Analysis

```{r clustering, exercise=TRUE}
# Hierarchical clustering
food_matrix <- as.matrix(food_imputed[, food_vars])
rownames(food_matrix) <- food_imputed$Country

# Calculate distance and perform clustering
dist_matrix <- dist(scale(food_matrix))
hc <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram
plot(hc, main = "Hierarchical Clustering of Countries",
     xlab = "Country", ylab = "Distance", sub = "")
rect.hclust(hc, k = 4, border = 2:5)
```

### Regional Comparisons

```{r regional-comparison, exercise=TRUE}
# Compare regions on key foods
regional_summary <- food_imputed %>%
  group_by(Geographic_Location) %>%
  summarise(
    Avg_Garlic = mean(Garlic, na.rm = TRUE),
    Avg_Olive_Oil = mean(Olive_Oil, na.rm = TRUE),
    Avg_Tea = mean(Tea, na.rm = TRUE),
    Avg_Coffee = mean(Ground_Coffee, na.rm = TRUE),
    Avg_Butter = mean(Butter, na.rm = TRUE),
    .groups = 'drop'
  )

print(regional_summary)

# Visualize
regional_long <- regional_summary %>%
  pivot_longer(cols = starts_with("Avg_"), 
               names_to = "Food", 
               values_to = "Average",
               names_prefix = "Avg_")

ggplot(regional_long, aes(x = Geographic_Location, y = Average, fill = Food)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Food Consumption by Region",
       x = "Geographic Region",
       y = "Average % Households") +
  scale_x_discrete(labels = c("C" = "Central", "N" = "Northern", 
                               "S" = "Southern", "U" = "UK/Ireland", 
                               "X" = "Benelux"))
```

## Summary and Key Takeaways

### What We've Learned

1. **Cultural Patterns**: Food consumption reflects clear geographic and cultural patterns across Europe
2. **Mediterranean vs Northern**: Garlic and olive oil distinguish Southern European countries
3. **UK Uniqueness**: England and Ireland show distinctive patterns with high tea consumption
4. **PCA Interpretation**: Multiple PCs reveal different aspects of variation:
   - PC1: Mediterranean vs Northern divide (Garlic, Olive oil)
   - PC2: Other consumption patterns
   - PC3: Tea drinking patterns (UK/Ireland)
5. **Data Quality**: Food consumption data has more variability than biological measurements

### Key Differences Identified

**Portugal and Spain vs Sweden and Denmark:**
- **Higher**: Garlic, Olive oil, Oranges
- **Lower**: Frozen vegetables, Tinned fruit, Jam
- **Cultural interpretation**: Mediterranean diet vs Nordic diet

### Practice Exercise

```{r final-exercise, exercise=TRUE}
# Your comprehensive analysis task:
# 1. Perform PCA on the food data
# 2. Identify which food items best distinguish Northern from Southern Europe
# 3. Create a visualization showing this distinction

# Start your analysis here:
pca_final <- prcomp(food_imputed[, food_vars], scale. = TRUE)

# Add your code below:

```

```{r final-exercise-solution}
# Complete analysis
pca_final <- prcomp(food_imputed[, food_vars], scale. = TRUE)

# Identify key variables
loadings_pc1 <- sort(pca_final$rotation[,1], decreasing = TRUE)
cat("Foods more common in one direction:\n")
print(head(loadings_pc1, 3))
cat("\nFoods more common in opposite direction:\n")
print(tail(loadings_pc1, 3))

# Visualize
scores_final <- data.frame(
  PC1 = pca_final$x[,1],
  PC2 = pca_final$x[,2],
  Country = food_imputed$Country,
  Region = food_imputed$Geographic_Location
)

ggplot(scores_final, aes(x = PC1, y = PC2, color = Region, label = Country)) +
  geom_point(size = 4) +
  geom_text(vjust = -0.7, size = 2.5) +
  labs(title = "European Food Consumption Patterns") +
  theme_minimal()
```

### Discussion Questions

1. **Why do you think there are such clear regional patterns in food consumption?**
   - Climate and agriculture
   - Cultural traditions
   - Historical trade patterns

2. **What other factors might explain the differences beyond geography?**
   - Economic development
   - Urbanization
   - Immigration patterns
   - EU integration effects

3. **How reliable is this data for making inferences about actual consumption?**
   - Based on household presence, not quantity
   - Survey-based (potential biases)
   - Temporal snapshot (may change over time)