---
title: "Multivariate Data Analysis: Food data set"
subtitle: "Principal Component Analysis with R"
author: "Rico Derks, Aswin Verhoeven"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "fos-2026-food"
  version: "1.0"
---

```{r setup, include=FALSE}
library(learnr)
# pcaMethods is loaded first on purpose: it masks a few base/stats functions,
# and loading the tidyverse packages afterwards restores the ones we need.
library(pcaMethods)
library(ggplot2)
library(dplyr)
library(plotly)
library(factoextra)
library(knitr)
library(DT)
library(tidyr)

# Set up tutorial options
tutorial_options(exercise.timelimit = 60)

# load the data
food_data <- read.csv(
  file = "./foods.csv",
  header = TRUE
)

# Store variable names for easy reference
food_vars <- names(food_data)[5:ncol(food_data)]
n_vars <- length(food_vars)

# Imputed version, used by all exercises from the PCA section onwards.
# (See the "Handling Missing Values" section for how this is done.)
food_imputed <- food_data
for (col in food_vars) {
  food_imputed[[col]][is.na(food_imputed[[col]])] <- mean(food_imputed[[col]],
                                                          na.rm = TRUE)
}

# Colours and labels for the geographic regions, used in several plots
region_colors <- c("C" = "#E69F00", "S" = "#D55E00", "N" = "#0072B2",
                   "U" = "#009E73", "X" = "#CC79A7")
region_labels <- c("C" = "Central", "S" = "Southern", "N" = "Northern",
                   "U" = "UK/Ireland", "X" = "Benelux")
```

```{r prep-pca}
# Setup chunk: makes the PCA model available to the exercises below.
# learnr runs each exercise in its own environment, so anything an exercise
# needs has to come from the global setup chunk or from a setup chunk like
# this one, referenced with the exercise.setup chunk option.
# (Note: a setup chunk must NOT be named "<something>-setup" unless an
# exercise with that exact name exists, so we use the "prep-" prefix.)
pca_food <- prcomp(food_imputed[, food_vars],
                   scale. = TRUE,
                   center = TRUE)

variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)
```

```{r prep-scores, exercise.setup="prep-pca"}
scores_df <- data.frame(
  PC1 = pca_food$x[, 1],
  PC2 = pca_food$x[, 2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location,
  Capitol_Latitude = food_imputed$Capitol_Latitude
)
```

## How to use this tutorial

This works exactly like the IRIS tutorial, so if you have just finished that one
you can go straight to the next section.

Everything runs in your browser; you cannot break anything. The grey boxes are
live R code: press **Run Code** to execute a box, **Start Over** to restore the
original code, and **Solution** (where it is offered) to see one possible answer.
You are encouraged to change the code and run it again.

Answer the questions with **Submit Answer** and read the explanation that
follows - it is usually the most useful part, also when you were right. Sections
appear one at a time; press **Continue** to move on, and your progress is
remembered if you close the page.

Remember that each code box runs independently: an object created in one box is
not automatically available in the next.

## Background

Data were collected to investigate the consumption pattern of a number of provisions in different European countries. The purpose of the investigation was to examine similarities and differences between the countries and the possible explanations.

### Objective

In this tutorial you will:

- Explain why there are groupings in multivariate plots
- Identify data characteristics that differentiate Portugal and Spain from Sweden and Denmark
- Understand geographic and cultural patterns in food consumption across Europe
- Practice interpreting PCA results with real-world data

### The Data

The dataset consists of **20 variables** (different foods) and **16 observations** (European countries). The values are the percentages of households in each country where a particular product was found.

Let's explore the data structure:

```{r explore-data, exercise=TRUE}
# Display the first few rows
head(food_data, 10)

# Get summary statistics for the food variables
summary(food_data[, food_vars])
```

```{r explore-data-solution}
head(food_data, 10)
summary(food_data[, food_vars])
```

### Data Exploration

```{r data-structure, exercise=TRUE}
# Check the structure
str(food_data)

# Count countries by geographic location
table(food_data$Geographic_Location)

# Geographic locations:
# C = Central, S = Southern, N = Northern, U = UK/Ireland, X = Benelux
```

<div id="data-structure-hint">
**Hint:** The `Geographic_Location` variable groups the countries:
C = Central Europe, S = Southern Europe, N = Northern Europe,
U = UK/Ireland, X = Benelux.
</div>

### Interactive Data Table

```{r data-table, exercise=TRUE}
# Create an interactive table to explore the data
datatable(food_data,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "European Food Consumption Data")
```

## Principal Component Analysis

### Handling Missing Values

Notice that some values are missing (NA). We need to address this before PCA:

```{r check-missing, exercise=TRUE}
# Check for missing values
colSums(is.na(food_data))

# Visualize missing data pattern
missing_pattern <- sapply(food_data[, food_vars], function(x) sum(is.na(x)))
barplot(missing_pattern, las = 2, main = "Missing Values per Variable",
        cex.names = 0.7, col = "steelblue")
```

```{r handle-missing, exercise=TRUE}
# Option 1: Remove rows with missing values (simple approach)
food_complete <- na.omit(food_data)
cat("Countries remaining:", nrow(food_complete), "\n")
cat("Countries removed:", food_data$Country[!complete.cases(food_data)])

# Option 2: Impute missing values with column means
food_imputed <- food_data
for(col in food_vars) {
  food_imputed[[col]][is.na(food_imputed[[col]])] <- mean(food_imputed[[col]], na.rm = TRUE)
}

# We'll use the imputed version for our analysis
cat("\nUsing imputed dataset with all", nrow(food_imputed), "countries")
```

#### Why SIMCA did not ask you to do this

If you ran this exercise in SIMCA, you never had to choose between deleting and
imputing: SIMCA simply kept all 16 countries. That is because SIMCA fits its PCA
models with the **NIPALS** algorithm, which works one component at a time and
can just skip the missing cells while it iterates. `prcomp()` cannot do that -
it needs a complete matrix - which is why we had to patch the holes first.

`pcaMethods` does give you NIPALS in R, so we can check whether our imputation
changed anything:

```{r nipals-compare, exercise=TRUE}
# Original data, missing values and all
X_raw <- as.matrix(food_data[, food_vars])
rownames(X_raw) <- food_data$Country

# NIPALS handles the NAs directly, like SIMCA does
pca_nipals <- pca(X_raw, method = "nipals", nPcs = 3,
                  scale = "uv", center = TRUE)

# Compare with the mean-imputed version
X_imp <- as.matrix(food_imputed[, food_vars])
pca_imp <- pca(X_imp, method = "svd", nPcs = 3, scale = "uv", center = TRUE)

data.frame(
  Component = paste0("PC", 1:3),
  R2cum_NIPALS = round(R2cum(pca_nipals), 3),
  R2cum_imputed = round(R2cum(pca_imp), 3)
)
```

The two are practically identical. With only 3 missing values out of 320, mean
imputation was harmless here - but that is a conclusion you should *check*
rather than assume. If a variable were missing for half of the countries,
replacing those values by the mean would quietly invent structure that is not
in the data.

### Performing PCA

```{r pca-basic, exercise=TRUE}
# Perform PCA on the food variables (columns 5 onwards)
# Using scale = TRUE for standardization (UV scaling)
pca_food <- prcomp(food_imputed[, food_vars], 
                   scale. = TRUE, 
                   center = TRUE)

# Display summary
summary(pca_food)

# Calculate variance explained
variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("\nVariance explained by first 3 PCs:\n")
cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("PC3:", round(r2_values[3]*100, 1), "%\n")
cat("\nCumulative variance (PC1+PC2):", round(r2_cumulative[2]*100, 1), "%\n")
cat("Cumulative variance (PC1+PC2+PC3):", round(r2_cumulative[3]*100, 1), "%\n")
```

```{r pca-basic-solution}
pca_food <- prcomp(food_imputed[, food_vars], scale. = TRUE, center = TRUE)
summary(pca_food)

variance_explained <- pca_food$sdev^2
total_variance <- sum(variance_explained)
r2_values <- variance_explained / total_variance
r2_cumulative <- cumsum(r2_values)

cat("PC1:", round(r2_values[1]*100, 1), "%\n")
cat("PC2:", round(r2_values[2]*100, 1), "%\n")
cat("Cumulative (PC1+PC2):", round(r2_cumulative[2]*100, 1), "%\n")
```

### Model quality: R² and Q²

The variance explained above is SIMCA's **R²cum**. On its own it is not enough
to judge a model, because R² can only ever increase when you add a component.
**Q²cum** is the cross-validated counterpart: the observations are split into 7
groups, each group is left out in turn, and the model built on the rest is used
to predict it. Q² therefore measures how much the model can *predict*, not just
describe - and unlike R², it can go **down** when a component starts fitting
noise.

```{r r2-q2, exercise=TRUE}
set.seed(2026)  # cross-validation splits the data randomly

X <- as.matrix(food_imputed[, food_vars])
rownames(X) <- food_imputed$Country

pca_cv <- pca(X, method = "svd", nPcs = 5, scale = "uv", center = TRUE)

r2cum <- R2cum(pca_cv)
q2cum <- Q2(pca_cv, X, fold = 7, verbose = FALSE)

data.frame(
  Component = paste0("PC", seq_along(r2cum)),
  R2cum = round(r2cum, 3),
  Q2cum = round(q2cum[seq_along(r2cum)], 3)
)
```

Look carefully at the two columns. R²cum climbs steadily, but Q²cum peaks
around the third component and then falls away. That is exactly why SIMCA's
"Autofit" stops at 3 components for this data set.

Note that with only 16 countries the cross-validation is noisy, so your Q²
values will shift a little every time the random split changes. The *pattern* -
low values throughout, peaking at 3 components - is stable; the exact numbers
are not.

### Quiz: Model Quality

```{r quiz-pca-quality, echo=FALSE}
quiz(
  question("Based on the PCA summary, how many principal components would you consider sufficient to describe the data?",
    answer("1 component"),
    answer("2 components"),
    answer("3 components", correct = TRUE),
    answer("All 20 components"),
    message = "PC1 explains 31.3% and PC2 19.2%, so the first two components together reach only 50.5%. PC3 adds another 13.7%, bringing the total to 64.2%. That is why we will also look at PC1 vs PC3 later in this tutorial: PC3 still contains real structure. SIMCA's 'Autofit' also selects 3 components for this data set."
  ),
  question("What can you say about the quality of this PCA model compared to the IRIS dataset?",
    answer("Much better - higher R²"),
    answer("About the same"),
    answer("Not as good - lower R²", correct = TRUE),
    answer("Cannot compare different datasets"),
    message = "For IRIS, two components already covered 95.8% of the variance; here two components reach only 50.5%. Food consumption data is more complex and variable than flower measurements, so we expect lower R² values. This is normal for behavioural/social data."
  ),
  question("Q²cum for this model never rises above about 0.2, and starts dropping after the third component. What does that mean?",
    answer("The model describes some real structure, but its predictive ability is poor", correct = TRUE),
    answer("The model is excellent, because Q² is positive"),
    answer("The model is broken and the results cannot be interpreted at all"),
    answer("Q² is irrelevant for PCA"),
    message = "Compare this with IRIS, where Q²cum reached 0.71. A Q² of around 0.15 means the model captures a genuine but weak pattern: you can use it to explore and to generate hypotheses about which countries resemble each other, but you should not expect it to predict a new country's shopping basket. The fact that Q² falls after the third component is the signal to stop adding components - the fourth and fifth are fitting noise.",
    correct = "Correct - and note that R² alone would never have told you this. R² keeps climbing all the way to component 5."
  )
)
```

## Scores Plot (PC1 vs PC2)

The scores plot shows how countries are positioned based on their food consumption patterns:

```{r scores-plot, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare scores data
scores_df <- data.frame(
  PC1 = pca_food$x[,1],
  PC2 = pca_food$x[,2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location,
  Capitol_Latitude = food_imputed$Capitol_Latitude
)

# Create scores plot colored by geographic location
ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, hjust = 0.5, size = 3) +
  labs(title = "PCA Scores Plot - European Food Consumption",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "% variance)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r scores-plot-solution}
scores_df <- data.frame(
  PC1 = pca_food$x[,1],
  PC2 = pca_food$x[,2],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location
)

ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, size = 3) +
  labs(title = "PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "%)")) +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal()
```

### Interactive Scores Plot

```{r interactive-scores, exercise=TRUE, exercise.setup="prep-scores"}
# Create interactive plot with plotly
p <- ggplot(scores_df, aes(x = PC1, y = PC2, color = Geographic_Location,
                           text = paste("Country:", Country))) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(aes(label = Country), vjust = -0.7, size = 3, show.legend = FALSE) +
  labs(title = "Interactive PCA Scores Plot",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC2 (", round(r2_values[2]*100, 1), "%)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal()

ggplotly(p, tooltip = c("text", "colour"))
```

### Quiz: Identifying Clusters

```{r quiz-clusters, echo=FALSE}
quiz(
  question("Are there clear clusters present in the scores plot?",
    answer("No clusters"),
    answer("Yes, some regional patterns", correct = TRUE),
    answer("Perfect separation by region"),
    answer("Only 2 distinct clusters"),
    message = "Colouring by geographic location reveals roughly three groups: the Northern countries (Sweden, Denmark, Norway, Finland), the Southern countries (Portugal, Italy, Spain, and Austria alongside them), and the rest. The groups overlap, so this is a tendency rather than a clean separation."
  ),
  question("Which two countries lie furthest apart in the PC1 vs PC2 scores plot?",
    answer("Portugal and Sweden", correct = TRUE),
    answer("Germany and France"),
    answer("England and Ireland"),
    answer("Norway and Denmark"),
    message = "Portugal sits at the far negative end of PC1 (-4.6) and Sweden at the far positive end of both PC1 (3.2) and PC2 (4.2): they are the two poles of the Mediterranean-versus-Nordic contrast. Note that England and Ireland are NOT extreme here (Ireland sits almost exactly at PC1 = 0). They only stand out once we look at PC3, later in this tutorial."
  )
)
```

## Outliers: Hotelling's T² and DModX

Before interpreting the loadings, check whether any single country is distorting
the model. **Hotelling's T²** finds strong outliers lying far from the centre
*within* the model plane, drawn as a confidence ellipse on the scores plot.
**DModX** finds observations lying far *away* from the plane, which a scores
plot cannot show you at all.

```{r outliers, exercise=TRUE, exercise.setup="prep-scores"}
n <- nrow(scores_df)
A <- 2
sd_pc <- apply(pca_food$x[, 1:A], 2, sd)

scores_df$T2 <- rowSums((pca_food$x[, 1:A] / rep(sd_pc, each = n))^2)
t2_crit <- (A * (n - 1) * (n + 1)) / (n * (n - A)) * qf(0.95, A, n - A)

theta <- seq(0, 2 * pi, length.out = 200)
ellipse <- data.frame(
  PC1 = sd_pc[1] * sqrt(t2_crit) * cos(theta),
  PC2 = sd_pc[2] * sqrt(t2_crit) * sin(theta)
)

ggplot(scores_df, aes(x = PC1, y = PC2)) +
  geom_path(data = ellipse, colour = "grey40", linetype = "dashed") +
  geom_point(aes(colour = Geographic_Location), size = 4, alpha = 0.8) +
  geom_text(aes(label = Country), vjust = -0.9, size = 3) +
  scale_color_manual(values = region_colors, labels = region_labels) +
  labs(title = "Scores plot with 95% Hotelling's T² ellipse",
       colour = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")

outside <- scores_df$Country[scores_df$T2 > t2_crit]
cat("Critical value (95%):", round(t2_crit, 2), "\n")
cat("Countries outside the ellipse:",
    if (length(outside) == 0) "none" else paste(outside, collapse = ", "), "\n")

# The two most extreme countries, even though they stay inside:
head(scores_df[order(-scores_df$T2), c("Country", "T2")], 3)
```

Nothing falls outside the ellipse. That is worth pausing on: Sweden (T² = 6.3)
and Portugal (T² = 3.7) are clearly the most extreme countries - the same two
poles you identified in the scores plot earlier - yet with only 16 observations
the 95% ellipse is so wide that neither crosses it. A small data set has very
little power to flag outliers, so the absence of a flag is not evidence that
everything is fine.

```{r dmodx, exercise=TRUE}
X <- as.matrix(food_imputed[, food_vars])
rownames(X) <- food_imputed$Country

pca_cv <- pca(X, method = "svd", nPcs = 3, scale = "uv", center = TRUE)

dmodx_df <- data.frame(
  Country = food_imputed$Country,
  DModX = as.numeric(DModX(pca_cv, X))
)

ggplot(dmodx_df, aes(x = reorder(Country, DModX), y = DModX)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_hline(yintercept = 2, linetype = "dotted", colour = "red") +
  coord_flip() +
  labs(title = "DModX per country",
       subtitle = "Dashed = model average (1), red dotted = rule-of-thumb limit (2)",
       x = NULL, y = "DModX (normalised)") +
  theme_minimal()
```

### Quiz: Outlier diagnostics

```{r quiz-outliers, echo=FALSE}
quiz(
  question("Based on the T² ellipse and the DModX plot, are there countries you should remove?",
    answer("No - nothing is extreme enough to justify removal", correct = TRUE),
    answer("Yes, Sweden and Portugal, because they are the most extreme"),
    answer("Yes, all Northern European countries"),
    answer("Yes, the countries with missing values"),
    message = "No country falls outside the 95% T² ellipse, and the highest DModX (Holland, 1.5) does not come close to twice the model average. Sweden and Portugal are extreme, but being extreme is not the same as being an outlier: they are extreme precisely because their diets differ most, which is the pattern we are trying to study. Removing them would delete the very signal you are after. Outlier diagnostics exist to make you go and LOOK at a sample, not to delete it automatically."
  ),
  question("Why is DModX worth looking at when you already have the scores plot?",
    answer("A country can sit comfortably inside the ellipse and still be poorly described by the model", correct = TRUE),
    answer("DModX is just another way of drawing the scores plot"),
    answer("DModX replaces the need for R² and Q²"),
    answer("DModX measures how important each food item is"),
    message = "The scores plot only shows you the part of an observation that the model DOES capture. DModX measures the residual - the part it misses. An observation can look perfectly ordinary in the scores plot and still have a large DModX, which is why SIMCA shows you both."
  )
)
```

## Loadings Plot

<div class="alert alert-info">
**A word of warning about signs.** The direction of a principal component is
arbitrary: a component and its mirror image describe exactly the same model. Your
plot may therefore be flipped compared with the SIMCA handout or with a fellow
student's screen, and the sign of a loading on its own means nothing.

What *is* meaningful is the relative arrangement: which foods sit together,
which sit on opposite sides of the origin, and how far each lies from the
centre. Always read a loadings plot together with its scores plot - foods in one
corner of the loadings plot are the ones consumed most in the countries sitting
in that same corner of the scores plot.
</div>

The loadings plot shows which food items contribute most to the patterns we see:

```{r loadings-plot, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare loadings data
loadings_df <- data.frame(
  Variable = rownames(pca_food$rotation),
  PC1 = pca_food$rotation[,1],
  PC2 = pca_food$rotation[,2]
)

# Create loadings plot
ggplot(loadings_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, color = "red", alpha = 0.7) +
  geom_text(aes(label = Variable), vjust = -0.5, hjust = 0.5, size = 2.5, check_overlap = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PCA Loadings Plot",
       x = "PC1 Loadings",
       y = "PC2 Loadings") +
  theme_minimal() +
  coord_fixed()
```

### Identify Top Contributing Variables

```{r top-loadings, exercise=TRUE, exercise.setup="prep-pca"}
# Find variables with strongest loadings on PC1
pc1_loadings <- sort(abs(pca_food$rotation[,1]), decreasing = TRUE)
cat("Top 5 variables contributing to PC1:\n")
print(head(pc1_loadings, 5))

# Find variables with strongest loadings on PC2
pc2_loadings <- sort(abs(pca_food$rotation[,2]), decreasing = TRUE)
cat("\nTop 5 variables contributing to PC2:\n")
print(head(pc2_loadings, 5))

# Show actual direction (positive or negative)
cat("\nPC1 loadings (with direction):\n")
print(sort(pca_food$rotation[,1], decreasing = TRUE)[1:5])
cat("\nPC1 loadings (negative direction):\n")
print(sort(pca_food$rotation[,1], decreasing = FALSE)[1:5])
```

### Quiz: Interpreting Loadings

```{r quiz-loadings, echo=FALSE}
quiz(
  question("What food items are causing the main clustering patterns in the scores plot?",
    answer("Garlic and olive oil at one end, crisp bread and frozen fish at the other", correct = TRUE),
    answer("Coffee and tea only"),
    answer("Butter and margarine only"),
    answer("All foods contribute equally"),
    message = "A loadings plot always has two ends, and both are informative. Garlic (-0.25) and olive oil (-0.15) have the strongest negative loadings on PC1 and pull the Mediterranean countries to that side. On PC2 the strongest positive loadings are crisp bread (0.39), frozen fish (0.37) and frozen vegetables (0.27), which is what sets the Northern countries apart. The further a variable lies from the origin, the more it contributes."
  )
)
```

## PC1 vs PC3 Analysis

Let's examine PC1 vs PC3 to understand other patterns:

```{r scores-pc1-pc3, exercise=TRUE, exercise.setup="prep-pca"}
# Prepare scores data for PC1 vs PC3
scores_df_pc3 <- data.frame(
  PC1 = pca_food$x[,1],
  PC3 = pca_food$x[,3],
  Country = food_imputed$Country,
  Geographic_Location = food_imputed$Geographic_Location
)

# Create scores plot
ggplot(scores_df_pc3, aes(x = PC1, y = PC3, color = Geographic_Location, label = Country)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(vjust = -0.7, hjust = 0.5, size = 3) +
  labs(title = "PCA Scores Plot - PC1 vs PC3",
       x = paste0("PC1 (", round(r2_values[1]*100, 1), "%)"),
       y = paste0("PC3 (", round(r2_values[3]*100, 1), "%)"),
       color = "Region") +
  scale_color_manual(values = region_colors,
                     labels = region_labels) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### PC3 Loadings

```{r loadings-pc3, exercise=TRUE, exercise.setup="prep-pca"}
# Examine PC3 loadings
loadings_pc3 <- data.frame(
  Variable = rownames(pca_food$rotation),
  PC1 = pca_food$rotation[,1],
  PC3 = pca_food$rotation[,3]
)

ggplot(loadings_pc3, aes(x = PC1, y = PC3)) +
  geom_point(size = 3, color = "red", alpha = 0.7) +
  geom_text(aes(label = Variable), vjust = -0.5, hjust = 0.5, size = 2.5, check_overlap = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "PCA Loadings Plot - PC1 vs PC3",
       x = "PC1 Loadings",
       y = "PC3 Loadings") +
  theme_minimal() +
  coord_fixed()

# Show top contributors to PC3
cat("\nTop contributors to PC3:\n")
print(sort(abs(pca_food$rotation[,3]), decreasing = TRUE)[1:5])
```

### Detailed Country Comparison

```{r country-detail, exercise=TRUE}
# Compare specific countries mentioned in the exercise
countries_of_interest <- c("England", "Ireland", "Portugal", "Spain", 
                           "Sweden", "Denmark")

# Extract their data
subset_data <- food_imputed %>%
  filter(Country %in% countries_of_interest) %>%
  select(Country, Garlic, Olive_Oil, Tea, Ground_Coffee)

print(subset_data)

# Visualize
subset_long <- subset_data %>%
  pivot_longer(cols = c(Garlic, Olive_Oil, Tea, Ground_Coffee),
               names_to = "Food", values_to = "Percentage")

ggplot(subset_long, aes(x = Country, y = Percentage, fill = Food)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Key Food Items - Selected Countries",
       y = "Percentage of Households")
```

### Quiz: PC1 vs PC3 Interpretation

```{r quiz-pc3, echo=FALSE}
quiz(
  question("England and Ireland are separated from the rest of Europe along PC3. Which variable contributes most to this separation?",
    answer("Their unusually LOW consumption of ground coffee", correct = TRUE),
    answer("Their high tea consumption"),
    answer("Their high tinned soup consumption"),
    answer("Their high yoghurt consumption"),
    message = "This one is worth looking at carefully. At first glance jam, tea and tinned soup seem to be the obvious explanation, but check the loadings: ground coffee has by far the largest PC3 loading (0.48). England (27%) and Ireland (30%) are far below every other country - the next lowest is Austria at 55%. Tea does not actually discriminate: England and Ireland are at 99%, but Holland (98%), Sweden (93%) and Denmark (92%) are just as high. Tinned soup does not either: England is at 76% but Ireland at 18%, which is exactly the European average. Jam does contribute (loading -0.31, both around 90% against a mean of 55%), just less than ground coffee."
  ),
  question("How does the Garlic and Olive oil intake of England/Ireland compare to the rest of Europe?",
    answer("Garlic much lower, olive oil about average", correct = TRUE),
    answer("Both much lower than average"),
    answer("Both much higher than average"),
    answer("Garlic about average, olive oil much lower"),
    message = "Garlic: Ireland (5%) is the lowest of all 16 countries and England (11%) is among the lowest, against a European mean of 42%. Olive oil is a different story: England is at 57% and Ireland at 31%, against a mean of 54% - so England is even slightly above average. Compare that with Spain (86% garlic, 91% olive oil) or Italy (80% and 94%). It is a useful reminder to check the actual variable plots instead of assuming that two variables which sit close together in one loadings plot always behave the same way."
  )
)
```

## Enhanced Visualizations

### Biplot - Scores and Loadings Together

```{r biplot, exercise=TRUE, exercise.setup="prep-pca"}
# Create a biplot showing both scores and loadings
fviz_pca_biplot(pca_food,
                label = "var",
                col.ind = food_imputed$Geographic_Location,
                palette = region_colors,
                addEllipses = FALSE,
                col.var = "black",
                repel = TRUE,
                legend.title = "Region",
                title = "PCA Biplot - Food Consumption Patterns") +
  theme_minimal()
```

### Contribution of Variables

```{r variable-contribution, exercise=TRUE, exercise.setup="prep-pca"}
# Visualize variable contributions to PC1 and PC2
# Contributions to PC1
fviz_contrib(pca_food, choice = "var", axes = 1, top = 10) +
  labs(title = "Top 10 Variables Contributing to PC1") +
  theme_minimal()
```

```{r variable-contribution-pc2, exercise=TRUE, exercise.setup="prep-pca"}
# Contributions to PC2
fviz_contrib(pca_food, choice = "var", axes = 2, top = 10) +
  labs(title = "Top 10 Variables Contributing to PC2") +
  theme_minimal()
```

## Alternative Univariate Approaches

If multivariate methods don't provide sufficient clarity, what alternative approaches could you use?

```{r quiz-univariate, echo=FALSE}
question("Which univariate approaches could you use here instead of PCA? (select all that apply)",
  answer("A one-way ANOVA per food item, comparing the geographic regions", correct = TRUE),
  answer("A t-test per food item when comparing just two groups of countries", correct = TRUE),
  answer("Correcting for multiple testing, because you run one test per food item", correct = TRUE),
  answer("Reporting whichever food has the largest PC1 loading, without any test"),
  type = "multiple",
  message = "Go variable by variable: a one-way ANOVA across the five regions, or a t-test if you reduce it to two groups (say Southern versus Northern). Because you would be running 20 tests - one per food - you must correct for multiple testing (Bonferroni, or Benjamini-Hochberg if you prefer to control the false discovery rate). What you lose compared with PCA is the correlation structure: univariate tests treat garlic and olive oil as two unrelated findings, whereas the loadings plot shows at a glance that they travel together."
)
```

The heatmap below is a visual version of that variable-by-variable comparison:

```{r univariate-viz, exercise=TRUE}
# Heatmap of food consumption by country
# Select a subset of interesting foods
foods_subset <- c("Garlic", "Olive_Oil", "Tea", "Ground_Coffee", 
                  "Butter", "Margarine", "Frozen_Vegetables")

heatmap_data <- food_imputed %>%
  select(Country, all_of(foods_subset))

# Convert to long format
heatmap_long <- heatmap_data %>%
  pivot_longer(cols = -Country, names_to = "Food", values_to = "Percentage")

# Create heatmap
ggplot(heatmap_long, aes(x = Food, y = Country, fill = Percentage)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", 
                       midpoint = 50) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Food Consumption by Country",
       fill = "% Households")
```

### Clustering Analysis

```{r clustering, exercise=TRUE}
# Hierarchical clustering
food_matrix <- as.matrix(food_imputed[, food_vars])
rownames(food_matrix) <- food_imputed$Country

# Calculate distance and perform clustering
dist_matrix <- dist(scale(food_matrix))
hc <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram
plot(hc, main = "Hierarchical Clustering of Countries",
     xlab = "Country", ylab = "Distance", sub = "")
rect.hclust(hc, k = 4, border = 2:5)
```

### Regional Comparisons

```{r regional-comparison, exercise=TRUE}
# Compare regions on key foods
regional_summary <- food_imputed %>%
  group_by(Geographic_Location) %>%
  summarise(
    Avg_Garlic = mean(Garlic, na.rm = TRUE),
    Avg_Olive_Oil = mean(Olive_Oil, na.rm = TRUE),
    Avg_Tea = mean(Tea, na.rm = TRUE),
    Avg_Coffee = mean(Ground_Coffee, na.rm = TRUE),
    Avg_Butter = mean(Butter, na.rm = TRUE),
    .groups = 'drop'
  )

print(regional_summary)

# Visualize
regional_long <- regional_summary %>%
  pivot_longer(cols = starts_with("Avg_"), 
               names_to = "Food", 
               values_to = "Average",
               names_prefix = "Avg_")

ggplot(regional_long, aes(x = Geographic_Location, y = Average, fill = Food)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Food Consumption by Region",
       x = "Geographic Region",
       y = "Average % Households") +
  scale_x_discrete(labels = c("C" = "Central", "N" = "Northern", 
                               "S" = "Southern", "U" = "UK/Ireland", 
                               "X" = "Benelux"))
```

## Summary and Key Takeaways

### What We've Learned

1. **Cultural Patterns**: Food consumption reflects clear geographic and cultural patterns across Europe
2. **Mediterranean vs Northern**: Garlic and olive oil distinguish Southern European countries
3. **UK Uniqueness**: England and Ireland stand apart mainly because of their very low ground coffee consumption, not because of tea
4. **PCA Interpretation**: Multiple PCs reveal different aspects of variation:
   - PC1 (31.3%): Mediterranean vs the rest (garlic, olive oil at the negative end)
   - PC2 (19.2%): the Nordic pattern (crisp bread, frozen fish, frozen vegetables)
   - PC3 (13.7%): England and Ireland (low ground coffee, high jam)
5. **Data Quality**: Food consumption data has more variability than biological measurements. R²cum reaches 0.64 at three components, but Q²cum peaks around 0.2 and then falls: the model describes a real but weak pattern, good enough to explore with, not good enough to predict with.
6. **Diagnostics**: no country is a genuine outlier here, but note *why* - with only 16 observations the Hotelling's T² ellipse is very wide, so a small data set has little power to detect outliers in the first place.
7. **Missing data**: SIMCA's NIPALS algorithm handles missing values directly, `prcomp()` does not. Mean imputation was harmless with 3 missing cells out of 320, but that needs checking rather than assuming.

### Key Differences Identified

**Portugal and Spain vs Sweden and Denmark** (average % of households, the four countries only):

- **Higher**: Garlic (88 vs 10), Olive oil (92 vs 39), Instant coffee (33 vs 15)
- **Lower**: Crisp bread (11 vs 64), Jam (27 vs 70), Frozen vegetables (5 vs 44), Tinned fruit (19 vs 52)
- **Cultural interpretation**: Mediterranean diet vs Nordic diet - fresh, oil- and garlic-based cooking against preserved, frozen and tinned products

### Practice Exercise

```{r final-exercise, exercise=TRUE}
# Your comprehensive analysis task:
# 1. Perform PCA on the food data
# 2. Identify which food items best distinguish Northern from Southern Europe
# 3. Create a visualization showing this distinction

# Start your analysis here:
pca_final <- prcomp(food_imputed[, food_vars], scale. = TRUE)

# Add your code below:

```

```{r final-exercise-solution}
# Complete analysis
pca_final <- prcomp(food_imputed[, food_vars], scale. = TRUE)

# Identify key variables
loadings_pc1 <- sort(pca_final$rotation[,1], decreasing = TRUE)
cat("Foods more common in one direction:\n")
print(head(loadings_pc1, 3))
cat("\nFoods more common in opposite direction:\n")
print(tail(loadings_pc1, 3))

# Visualize
scores_final <- data.frame(
  PC1 = pca_final$x[,1],
  PC2 = pca_final$x[,2],
  Country = food_imputed$Country,
  Region = food_imputed$Geographic_Location
)

ggplot(scores_final, aes(x = PC1, y = PC2, color = Region, label = Country)) +
  geom_point(size = 4) +
  geom_text(vjust = -0.7, size = 2.5) +
  labs(title = "European Food Consumption Patterns") +
  theme_minimal()
```

### Discussion Questions

1. **Why do you think there are such clear regional patterns in food consumption?**
   - Climate and agriculture
   - Cultural traditions
   - Historical trade patterns

2. **What other factors might explain the differences beyond geography?**
   - Economic development
   - Urbanization
   - Immigration patterns
   - EU integration effects

3. **How reliable is this data for making inferences about actual consumption?**
   - Based on household presence, not quantity
   - Survey-based (potential biases)
   - Temporal snapshot (may change over time)